A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-05 latex/main.tex
Minor tweaks to results sections

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that incrementally adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts~\cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that this ARG is highly congruent with the UShER phylogeny built 
from the same dataset as well as the Pango nomenclature system~\cite{Rambaut2020},
and also accurately captures evolutionary features such as mutational 
spectra~\cite{Bloom2023} and deletions\cite{Li2023indels}.
We show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated in a community 
process, and provides novel insights into the relationships between these
recombinant lineages. We compare the recombination events detected by 
sc2ts with the state-of-the-art combination of UShER and 
RIPPLES~\cite{Turakhia2022}, showing that sc2ts is more sensitive,
detecting more of the well-characterized recombination events 
curated by the Pango lineage designation community.
The output of sc2ts---incorporating
genealogical relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit~\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
These resources
will help us to learn from the vast troves of data collected during the
pandemic, and to prepare for the next.

@@

[FIXME Going in too fast here]
Sc2ts reconstructs an ARG by incrementally adding
daily batches of sequences (Figure~\ref{fig:methodology}).
First, sc2ts infers likely ``copying paths'' connecting each sample
in the batch to node(s) in the current ARG
using a simplified version of the Li and Stephens (LS) model~\cite{LiStephens2003}.
% Important to go into a bit of detail about LS here as it's so fundamental
% and novel in the area.
The LS model is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination
on genetic inheritance~\cite{Delaneau2019-wl,Browning2018-nk},
typically parameterized by population-scaled mutation and
recombination rates varying along the genome~\cite{Donnelly2010-coalescent}.
In the interest of interpretability, we simplified this model
to use a single parameter $k$
that controls the number of recurrent mutations
that is allowed before switching to a different parent via recombination
% This is quite a big thing really, and ultimately what determines
% the number of recombinants detected.
(we use $k=4$ here; see STAR Methods and Discussion).
We used the highly efficient LS HMM implementation in tsinfer~\cite{Kelleher2019},
with some additional optimizations enabled by our simplified model,
to find exact Viterbi solutions of these copying paths for each sample
given the current ARG.
% TODO note the parsimony is an OK model here, as Sally suggests?
% Could cite paper showing parsimony works well too.
Note that this is a form of parsimony, 
with the recombination penalty $k$
defining the relative parsimoniousness of
recombinant versus non-recombinant solutions to the HMM.
The per-sample copying paths produced by the HMM
(most of which do not involve recombination)
form clusters of samples within a batch,
which are then resolved using standard phylogenetic techniques.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced
by the inherent greediness of this strategy.
Sc2ts is implemented in Python,
building on core PyData infrastructure~\cite{Harris2020array,Virtanen2020scipy,
Mckinney2010data,Lam2015numba} and
bioinformatics packages~\cite{Shirley2015efficient,Kunzmann2018biotite,Kunzmann2023biotite}.
See STAR Methods for details on all aspects of the method.

@@

Deletions are an important part of SARS-CoV-2 evolution~\cite{Jeronimo2023}
but are not incorporated into UShER trees and
therefore must be accounted for separately.
Because deletions are difficult to distinguish
% SHZ: Sequencing errors or misleading alignment.
from bioinformatic errors
and their multi-base nature presents challenges
to the Markovian structure of the LS HMM,
we masked out all gap characters as missing data during primary inference.
To approximate the effects of the most important deletions into the ARG,
we remapped data using \textit{post hoc} parsimony at 75 sites
involved in deletions with a frequency of
$\geq 1\%$ reported by Li et al.~\cite{Li2023indels}
during post-processing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known
high-frequency deletions (Table~\ref{tab:major_dels}) and
recurrent deletions (Table~\ref{tab:rec_dels}).
For example,
the three defining deletions of Alpha~\cite{RambautAlpha2020}
(11288:9, 21765:6, 21991:3)
are correctly mapped above the Alpha origin in the ARG
(Figure~\ref{fig:subgraph_alpha}).
The ARG captures known recurrent deletions
in the N-terminal domain of the S1 unit of Spike
(which is a deletion hot spot~\cite{McCarthy2021}),
for example,
the 21765:6 deletion ($\Delta$H69/V70),
which causes S-Gene Target Failure \cite{Walker2021} and
is associated with increased efficiency in cell entry~\cite{Meng2021}, and
the 21991:3 deletion
which is associated with antibody escape~\cite{McCarthy2021}.
See Document~\ref{sec:supplement_dels} for further analysis.

@@

The final class of events (Type IV) in Table~\ref{tab:pango_x_lineages}
corresponds to nodes that are not associated with a recombination event
in the ARG.
These vary in their plausibility of being non-recombinant.
For example, XN (Figure~\ref{fig:subgraphs}C) is only one mutation
distant from a BA.2 sample and has little additional evidence
to suggest recombination (Document~\ref{sec:dd_xn}).
XB, on the other hand,
has been thoroughly characterized~\cite{Gutierrez2022} and
may have misclassified by sc2ts due to limitations in
handling deletions or matching against long branches (Document \ref{sec:dd_xb}).

@@

To compare the performance of sc2ts and UShER+RIPPLES 
in terms of recombinant detection, we ran RIPPLES on the Viridian UShER phylogeny
subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
Note that we are interested in the direct output of sc2ts and UShER+RIPPLES
here, and do not perform QC filtering of either.
Using the RIPPLES parameter $p=4$ (matching $k=4$ used in the sc2ts ARG) 
resulted in 1174 reported recombination events.
We ran 3SEQ on each of the candidate trios (STAR Methods) and 601 
were classified as recombinant, giving an overall validation rate of 51\%.
Aligning the events reported by both methods using the sets of descendant
samples (STAR Methods), 468 of the events are shared by both,
with the intersection comprising 40\% of the RIPPLES events
and 55\% of the sc2ts events (Figure~\ref{fig:ripples_sc2ts_events}). 
The 706 recombination events output by RIPPLES that do not 
correspond to a recombination event in the sc2ts ARG can largely
be explained by a combination the choice of sites 
used for inference, and the strict time ordering imposed
by sc2ts (Document~\ref{sec:detailed_ripples}).
The 387 events detected by sc2ts and not by UShER+RIPPLES
would require a detailed dissection of the UShER tree
to explain, which is beyond the scope of this work.
However, the set does contain some well characterized events 
associated with the Pango X lineages.
Table~\ref{tab:ripples_x_lineages} shows the events detected 
by UShER+RIPPLES associated with samples assigned to Pango X
lineages. At the RIPPLES ``parsimony improvement'' parameter
$p=4$ (equivalent to $k=4$ used to build the sc2ts ARG)
only 9 of the Pango X lineages are present and
with unequivocal recombinant origin for 6 
(contrast with 16 Type I events in Table~\ref{tab:pango_x_lineages}).
We therefore reran RIPPLES with $p=3$, resulting in 4111 events overall,
and increasing the intersection of events with sc2ts to 595 (70\%;
Figure~\ref{fig:ripples_sc2ts_events}).
At this more lenient threshold, 23 of the 44 Pango X lineages
are present in the dataset, with unequivocal origins for 10 lineages, and
7 lineages associated with two competing events 
(Document~\ref{sec:detailed_ripples}).
XB and XAJ are among the lineages detected by UShER+RIPPLES,
strengthening evidence that these are false negatives 
for sc2ts (Table~\ref{tab:pango_x_lineages}).

>>>>

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that incrementally adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts~\cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that this ARG is highly congruent with the UShER phylogeny built 
from the same dataset as well as the Pango nomenclature system~\cite{Rambaut2020},
and also accurately captures evolutionary features such as mutational 
spectra~\cite{Bloom2023} and deletions\cite{Li2023indels}.
We show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated in a community 
process, and provides novel insights into the relationships between these
recombinant lineages. We compare the recombination events detected by 
sc2ts with the state-of-the-art combination of UShER and 
RIPPLES~\cite{Turakhia2022}, showing that sc2ts is more sensitive,
detecting more of the well-characterized recombination events 
curated by the Pango lineage designation community.
The output of sc2ts---incorporating
genealogical relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit~\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
These resources provide a strong platform for future research,
helping us to learn from the vast troves of data collected during the
pandemic, and to prepare for the next.

@@

Sc2ts is a real-time inference method, updated incrementally 
with daily batches of sequence data (Figure~\ref{fig:methodology}).
For each batch, sc2ts first infers likely ``copying paths'' connecting each sample
to node(s) in the current ARG
using a simplified version of the Li and Stephens (LS) model~\cite{LiStephens2003}.
% Important to go into a bit of detail about LS here as it's so fundamental
% and novel in the area.
The LS model is a Hidden Markov Model (HMM) widely used in human genomics
to approximate the effects of mutation and recombination
on genetic inheritance~\cite{Delaneau2019-wl,Browning2018-nk},
typically parameterized by population-scaled mutation and
recombination rates varying along the genome~\cite{Donnelly2010-coalescent}.
In the interest of interpretability, we simplified this model
to use a single parameter $k$
that controls the number of recurrent mutations
that is allowed before switching to a different parent via recombination
% This is quite a big thing really, and ultimately what determines
% the number of recombinants detected.
(we use $k=4$ here; see STAR Methods).
We used the highly efficient LS HMM implementation in tsinfer~\cite{Kelleher2019},
with some additional optimizations enabled by our simplified model,
to find exact Viterbi solutions of these copying paths for each sample
given the current ARG.
% TODO note the parsimony is an OK model here, as Sally suggests?
% Could cite paper showing parsimony works well too.
Note that this is a form of parsimony, 
with the recombination penalty $k$
defining the relative parsimoniousness of
recombinant versus non-recombinant solutions to the HMM.
The per-sample copying paths produced by the HMM
(most of which do not involve recombination)
form clusters of samples within a batch,
which are then resolved using standard phylogenetic techniques.
Finally, sc2ts attaches the phylogenies of the clusters to the current ARG, and
applies parsimony-improving heuristics to address issues introduced
by the inherent greediness of this strategy.
Sc2ts is implemented in Python,
building on core PyData infrastructure~\cite{Harris2020array,Virtanen2020scipy,
Mckinney2010data,Lam2015numba} and
bioinformatics packages~\cite{Shirley2015efficient,Kunzmann2018biotite,Kunzmann2023biotite}.
See STAR Methods for details on all aspects of the method.

@@

Deletions are an important part of SARS-CoV-2 evolution~\cite{Jeronimo2023}
but are not incorporated into UShER trees and
therefore must be accounted for separately.
Because deletions are difficult to distinguish
% SHZ: Sequencing errors or misleading alignment.
from bioinformatic errors
and their multi-base nature presents challenges
to the LS HMM,
we masked out all gap characters as missing data during primary inference.
To approximate the effects of the most important deletions into the ARG,
we remapped data using post hoc parsimony at 75 sites
involved in deletions with a frequency of
$\geq 1\%$ reported by Li et al.~\cite{Li2023indels}
during post-processing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known
high-frequency deletions (Table~\ref{tab:major_dels}) and
recurrent deletions (Table~\ref{tab:rec_dels}).
For example,
the three defining deletions of Alpha~\cite{RambautAlpha2020}
(11288:9, 21765:6, 21991:3)
are correctly mapped above the Alpha origin in the ARG
(Figure~\ref{fig:subgraph_alpha}).
The ARG captures known recurrent deletions
in the N-terminal domain of the S1 unit of Spike
(which is a deletion hot spot~\cite{McCarthy2021}),
for example,
the 21765:6 deletion ($\Delta$H69/V70),
which causes S-Gene Target Failure \cite{Walker2021} and
is associated with increased efficiency in cell entry~\cite{Meng2021}, and
the 21991:3 deletion
which is associated with antibody escape~\cite{McCarthy2021}.
See Document~\ref{sec:supplement_dels} for further analysis.

@@

The final class of events (Type IV) in Table~\ref{tab:pango_x_lineages}
corresponds to nodes that are not associated with a recombination event
in the ARG.
These vary in their plausibility of being non-recombinant.
For example, XN (Figure~\ref{fig:subgraphs}C) is only one mutation
distant from a BA.2 sample and has little additional evidence
to suggest recombination (Document~\ref{sec:dd_xn}).
XB, on the other hand,
has been thoroughly characterized~\cite{Gutierrez2022} and
is likely misclassified by sc2ts because the required
parent lineages were not present (Document \ref{sec:dd_xb}).

@@

To compare the performance of sc2ts and UShER+RIPPLES 
in terms of recombinant detection, we ran RIPPLES on the Viridian UShER phylogeny
subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
Note that we are interested in the direct output of sc2ts and UShER+RIPPLES
here, and do not perform QC filtering of either.
Using the RIPPLES parameter $p=4$ (matching $k=4$ used in the sc2ts ARG) 
resulted in 1174 reported recombination events.
We ran 3SEQ on each of the candidate trios (STAR Methods) and 601 
were classified as recombinant, giving an overall validation rate of 51\%.
Aligning the events reported by both methods using the sets of descendant
samples (STAR Methods), 468 of the events are shared by both,
with the intersection comprising 40\% of the RIPPLES events
and 55\% of the sc2ts events (Figure~\ref{fig:ripples_sc2ts_events}). 
The 706 recombination events output by RIPPLES that do not 
correspond to a recombination event in the sc2ts ARG can largely
be explained by a combination of the choice of sites 
used for inference, and the strict time ordering imposed
by sc2ts (Document~\ref{sec:detailed_ripples}).
The 387 events detected by sc2ts and not by UShER+RIPPLES
would require a detailed dissection of the UShER tree
to explain, which is beyond the scope of this work.
However, the set does contain some well characterized events 
associated with the Pango X lineages.
Table~\ref{tab:ripples_x_lineages} shows the events detected 
by UShER+RIPPLES associated with samples assigned to Pango X
lineages. At the RIPPLES ``parsimony improvement'' parameter
$p=4$ (equivalent to $k=4$ used to build the sc2ts ARG)
only 9 of the Pango X lineages are present and
with unequivocal recombinant origin for 6 
(contrast with 16 Type I events in Table~\ref{tab:pango_x_lineages}).
We therefore reran RIPPLES with $p=3$, resulting in 4111 events overall,
and increasing the intersection of events with sc2ts to 595 (70\%;
Figure~\ref{fig:ripples_sc2ts_events}).
At this more lenient threshold, 23 of the 44 Pango X lineages
are present in the dataset, with unequivocal origins for 10 lineages, and
7 lineages associated with two competing events 
(Document~\ref{sec:detailed_ripples}).
XB and XAJ are among the lineages detected by UShER+RIPPLES,
strengthening evidence that these are false negatives 
for sc2ts (Table~\ref{tab:pango_x_lineages}).
